---
title: "NBA Predictive Analytics"
author: "Bill Xiao"
date: "6/10/2023"
output:
  pdf_document:
  keep_tex: true
  pandoc_args: ["--quiet"]
  
---

## 0. Before your start
This file includes some R codes that are used in the project "Win-Loss Prediction in NBA Games".
The code is for eductional purpose only. It is provided 'as is' without warranty of any kind, either express or implied, warranties of fitness for a purpose, or the warranty of non-infringement. 
Although the authors try their best to test the tool, they make no warranty that
(1) it will meet your requirements;
(2) it will be secure or error-free;
(3) the results that may be obtained from the use of the code will be effective, accurate or reliable;
(4) any errors in the tool will be corrected.

The author assume no responsibility for errors or omissions in the code or related documentation. In no event shall the sponsor and the authors be liable to youor any third parties for any damages of any kind arising out of or in connection with the use of the code. 

R is a open-source statistical software and can be downloaded at www.r-project.org
The codes have been tested using R4.2.3

Install packages: 
```{r}
library(ada)
library(adabag)
library(car)
library(caret)
library(class)
library(corrplot)
library(dplyr)
library(e1071)
library(ggplot2)
library(glmnet)
library(hoopR)
library(knitr)
library(MASS)
library(partykit)
library(pROC)
library(randomForest)
library(rpart)
library(rpart.plot)
library("SmartEDA")
library(tictoc)
library(xfun)
library(zoo)

```


## 1. Extract Data
NBA team box scores (2019-2022) 
```{r}

#change this to your local directory where you stored the data file. All result files will be populated to this directory as well

setwd("C:/Users/17329/Desktop/6609_Projects/Codes")

for (i in 2019:2022){
 ds <- load_nba_team_box(i)
 write.csv(ds, paste0(i,"team_data.csv"),
           row.names = FALSE)}

```

## 2. Combine Data
```{r}
raw <- read.csv("2019team_data.csv")
for (i in 2020:2022){
  y <- paste0(i,"team_data.csv")
  data <- read.csv(y)
  raw <- bind_rows(raw,data)
  }
```

## 3. Add Variables
Two points field made. 
```{r}
raw$two_point_field_goals_made<-(raw$field_goals_made-raw$three_point_field_goals_made)
raw$two_point_field_goals_attempted<-(raw$field_goals_attempted-raw$three_point_field_goals_attempted)
raw$two_point_field_goal_pct <-
  round(100*raw$two_point_field_goals_made/raw$two_point_field_goals_attempted,1)

```

## 4. Variables of interest
```{r}
columns<-c("season","season_type","game_date","team_name","team_home_away",
           "team_score","team_winner","assists","blocks","steals",
           "defensive_rebounds","offensive_rebounds",
           "points_in_paint","fast_break_points",
           "total_turnovers","turnover_points",
           "free_throw_pct","free_throws_made","free_throws_attempted",
           "field_goal_pct","field_goals_made","field_goals_attempted",
           "two_point_field_goal_pct","two_point_field_goals_made",
           "two_point_field_goals_attempted",
           "three_point_field_goal_pct","three_point_field_goals_made",
           "three_point_field_goals_attempted",
           "flagrant_fouls","fouls","technical_fouls")

raw1 <- raw[,columns]
# (all(columns %in% colnames(raw)))

# Add a new column to identify each game
raw1$game_id <- rep(1:(nrow(raw1)/2), each = 2)

# Split the data frame into two: one for Team A and one for Team B
team_a <- raw1[seq(1, nrow(raw1), by = 2), ]
team_b <- raw1[seq(2, nrow(raw1), by = 2), ]

# Rename the columns in the Team B data frame to indicate that they are for the opponent
names(team_b) <- paste0("opponent_", names(team_b))
team_b <- rename(team_b, game_id=opponent_game_id)
# Merge the two data frames back together by game_id
merged <- left_join(team_a, team_b, by = "game_id")
nba <- merged[,c(c(1:17),20,23,26,30,36,38,c(40:49),52,55,58,62)]
nba$score_difference<-nba$team_score-nba$opponent_team_score
# write.csv(nba, "nba.csv")
nba$team_winner <- 1*(nba$team_winner==TRUE)
nba$team_home_away <- 1*(nba$team_home_away=="home")
nba$game_date <- as.Date(nba$game_date, format = "%m/%d/%Y")

# recoding variables
nba$Season <- nba$season
nba$Type <- nba$season_type
nba$Date <- nba$game_date
nba$Team <- nba$team_name
nba$Opponent <- nba$opponent_team_name
nba$HA <- nba$team_home_away
nba$WL <- nba$team_winner
nba$DIFF <- nba$team_score-nba$opponent_team_score
nba$AST <- nba$assists-nba$opponent_assists
nba$BLK <- nba$blocks-nba$opponent_blocks
nba$STL <- nba$steals-nba$opponent_steals
nba$DRB <- nba$defensive_rebounds-nba$opponent_defensive_rebounds
nba$ORB <- nba$offensive_rebounds-nba$opponent_offensive_rebounds
nba$PIP <- nba$points_in_paint-nba$opponent_points_in_paint
nba$FBP <- nba$fast_break_points-nba$opponent_fast_break_points
nba$TOV <- nba$total_turnovers-nba$opponent_total_turnovers
nba$TOVP <- nba$turnover_points-nba$opponent_turnover_points
nba$FTp <- nba$free_throw_pct-nba$opponent_free_throw_pct
nba$FGp <- nba$field_goal_pct-nba$opponent_field_goal_pct
nba$P2p <- nba$two_point_field_goal_pct-nba$opponent_two_point_field_goal_pct
nba$P3p <- nba$three_point_field_goal_pct-nba$opponent_three_point_field_goal_pct
nba$FOUL <- nba$fouls-nba$opponent_fouls

data <- nba[, c("Season", "Type", "Date", "Team", "Opponent", 
                "WL", "DIFF", "AST", "BLK", "STL", "DRB", 
                "ORB", "PIP", "FBP", "FOUL", "TOV", "TOVP", 
                "FTp", "FGp", "P2p", "P3p")]

```

## 5. Remove outlier
```{r}
data$Date <- as.Date(data$Date, format = "%m/%d/%Y")
data <- data[order(data$Date),]
data <- filter(data, !(Team %in% c("World", "Team LeBron", "Team Durant")))
data <- filter(data, !(Team %in% c("USA", "Team Giannis", 
                                   "Team Durant","Team LeBron")))

```

## 6. EDA
```{r}
# focus on non-standardized "nba" data set
eda<-nba[,39:60]
# Variables to use in later models
ExpData(eda,type=2)
# distribution of categorical variable
ExpCatViz(eda)
# distribution of numeric variable
ExpNumStat(eda[,8:22],Outlier = TRUE)

# Calculate the correlation matrix for numerical variables

correlation_matrix<-cor(eda[,8:22])
# Plot the correlation heatmap
# round(correlation_matrix,3)
# all correlations between predictors are below 0.8, which suggests that there is no strong linear relationship or multicollinearity among the predictors.
```

## 7. 3-game simple moving average
```{r}
# Scale numerical variables
numerical_vars <- c("DIFF", "AST", "BLK", "STL", "DRB", "ORB", 
                    "PIP", "FBP", "FOUL","TOV", "TOVP", "FTp", 
                    "FGp", "P2p", "P3p")  
data[numerical_vars] <- lapply(data[numerical_vars], scale)
# Specify categorical variables
categorical_vars <- c("Type", "Team", "Opponent", "WL")  
data[categorical_vars] <- lapply(data[categorical_vars], as.factor)

# Specify the variables for which to calculate the moving average
variables <- c("DIFF", "AST", "BLK", "STL", "DRB", "ORB", 
               "PIP", "FBP", "FOUL","TOV", "TOVP", "FTp", 
               "FGp", "P2p", "P3p")

# Calculate the three-game moving average with a lag of 1 for each team and each variable
data1 <- data %>%
  arrange(Team, Date) %>%
  group_by(Team) %>%
  mutate(across(all_of(variables), 
                ~ lag(rollmean(., 3, fill = NA, align = "right"), 1)))

# sum(is.na(data1)) # 0
data1 <- na.omit(data1)

train = data1[data1$Season!=2022, ]
test  = data1[data1$Season==2022, ]

# sum(data1$Season==2022) # 1317
# sum(data1$Season!=2022) # 3524

# evaluation metrics
metrics = function(CM){
  # CM = confusionMatrix
  TN=CM$table[1,1]
  FN=CM$table[1,2]
  FP=CM$table[2,1]
  TP=CM$table[2,2]
  recall=TP/(TP+FN)
  precision = TP/(TP+FP)
  accuracy=(TP+TN)/(sum(CM$table))
  F1 = 2*recall*precision/(recall+precision)
  result = data.frame(
    Metrics = c("Accuracy","F measure"),
    Values = c(accuracy, F1)
  )
  return(result)
}

# CM = confusionMatrix(as.factor(test$LR_WL), test$WL)

# CM = confusionMatrix(test$RFpredicted_WL, test$WL)

# metrics(CM)

```

## 8. Naive bayes
```{r}
# Specify the predictors
predictors <- c("Team", "Opponent", "DIFF", "AST", "BLK",
                "STL", "DRB", "ORB", "PIP", "FBP", "FOUL",
                "TOV", "TOVP", "FTp", "FGp", "P2p", "P3p")

# Fit a Naive Bayes model
NBmodel <- naiveBayes(WL ~ ., data = train[, c("WL", predictors)])

# print(NBmodel)

# Predict on the training set
train$predicted_WL <- predict(NBmodel, newdata = train[, predictors])

# Print the confusion matrix for the training set
# print(confusionMatrix(train$predicted_WL, train$WL))

# Predict on the test set
test$NBpredicted_WL <- predict(NBmodel, newdata = test[, predictors])

# Print the confusion matrix for the test set

# print(confusionMatrix(test$NBpredicted_WL, test$WL))

```

## 9. KNN
```{r}
# Find the optimal value of K using cross-validation
set.seed(1)
trainControl <- trainControl(method = "cv", number = 5) # 5-fold

KNNpredictors <- c("DIFF", "AST", "BLK", "STL", "DRB", "ORB", "PIP", "FBP", "FOUL","TOV", "TOVP", "FTp", "FGp", "P2p", "P3p")
# only include numeric variables in KNN setting

knnFit <- train(WL ~ ., data = train, 
                method = "knn", 
                tuneLength = 25, trControl = trainControl)
print(knnFit)

train_KNN_accuracy = 0.5528

# Create the plot
ggplot(data = knnFit$results, aes(x = k, y = Accuracy)) +
  geom_line() +
  geom_point() +
  labs(x = "Number of Neighbors (k)", y = "Cross-Validated Accuracy") +
  ggtitle("K-Nearest Neighbors Model Tuning") +
  theme_minimal()

# Predict on the test set using the optimal value of K
test$KNNpredicted_WL <- knn(train = train[, KNNpredictors], 
                             test = test[, KNNpredictors], 
                             cl = train$WL, k = knnFit$bestTune$k)

# Print the confusion matrix for the test set
print(confusionMatrix(test$KNNpredicted_WL, test$WL))
test_KNN_accuracy = 0.5254

```

## 10. Logistic regression 
```{r}
# Specify the predictors
LRpredictors <- c("Team", "Opponent", "DIFF", "AST", "BLK", "STL", "DRB", "ORB", "PIP", "FBP", "FOUL","TOV", "TOVP", "FTp", "FGp", "P2p", "P3p")

# Fit a logistic regression model
full_model <- glm(WL ~ ., train[, c("WL", LRpredictors)], 
                  family = binomial(link = "logit"))

print(summary(full_model))

# Perform backward selection
bw_model <- stepAIC(full_model, direction = "backward")

# Print the selected model
print(bw_model)

# Predict on the training set
train$LRpredicted_WL <- predict(bw_model, newdata = train[, LRpredictors], 
                                type = "response")

# Find the optimal probability threshold for the training set
roc_obj <- roc(train$WL, train$LRpredicted_WL)
coords_obj <- coords(roc_obj, "best")
optimal_threshold <- coords_obj["threshold"]
# print(optimal_threshold)

# Convert ROC object to a data frame for ggplot
roc_df <- data.frame(
  TPR = roc_obj$sensitivities,
  FPR = 1 - roc_obj$specificities,
  Thresholds = roc_obj$thresholds
)

# Create the ROC curve plot
ggplot(data = roc_df, aes(x = FPR, y = TPR)) +
  geom_line() +
  geom_point(data = roc_df[roc_df$Thresholds == optimal_threshold, ], color = "red") +
  labs(
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    title = "ROC Curve with Optimal Threshold"
  ) +
  theme_minimal() +
  annotate("text", x = .75, y = .45, 
           label = paste("Optimal Threshold: ",
                         round(optimal_threshold, 4)), 
           color = "red")

# Convert probabilities to class labels based on the optimal threshold
train$LR_WL <- ifelse(train$LRpredicted_WL > as.numeric(optimal_threshold), 1, 0)

# Print the confusion matrix for the training set
# print(confusionMatrix(as.factor(train$LR_WL), train$WL))

# Predict on the test set
test$LRpredicted_WL <- predict(bw_model, newdata = test[, LRpredictors], 
                             type = "response")

test$LR_WL <- ifelse(test$LRpredicted_WL > as.numeric(optimal_threshold), 1, 0)

# Print the confusion matrix for the test set
# print(confusionMatrix(as.factor(test$LR_WL), test$WL))

vif(bw_model)

```

## 11. Decision tree
```{r}
# Specify the predictors
DTpredictors <- c("Team", "Opponent", "DIFF", "AST", 
                  "BLK", "STL", "DRB", "ORB", "PIP", "FBP",
                  "FOUL","TOV", "TOVP", "FTp", "FGp", "P2p", "P3p")

# Fit a decision tree model
tree_model <- rpart(WL ~ ., data = train[, c("WL", LRpredictors)],
                    method = "class", 
                    control = rpart.control(cp = 0.01))

# Perform 5-fold cross-validation to determine the optimal tree size
set.seed(100)
cv_DTmodel <- train(WL ~ ., data = train[, c("WL", LRpredictors)],
                  method = "rpart", 
                  trControl = trainControl(method = "cv", number = 5))

# Print the cross-validation results
# print(cv_DTmodel)

# Get the optimal tree size (complexity parameter)
optimal_cp <- cv_DTmodel$bestTune$cp
print(paste("Optimal tree size (cp):", optimal_cp))

# Fit the final decision tree model with the optimal tree size
final_DT_model <- rpart(WL ~ ., data = train[, c("WL", LRpredictors)],
                        method = "class", 
                        control = rpart.control(cp = optimal_cp))

# Predict on the training set
train$DTpredicted_WL <- predict(final_DT_model, 
                                newdata = train[, c("WL", LRpredictors)], 
                                type = "class")

# Print the confusion matrix for the training set
# confusionMatrix(train$DTpredicted_WL, train$WL)

# Predict on the test set
test$DTpredicted_WL <- predict(final_DT_model, newdata = test, type = "class")

# Print the confusion matrix for the test set
# print(confusionMatrix(test$DTpredicted_WL, test$WL))

# png(filename = "tree.png", width = 1200, height = 500)
# rpart.plot(final_DT_model, yesno=2, type=4, extra=2)
# dev.off()

```

## 12. Random forest
```{r}
# Specify the predictors
RFpredictors <- c("Team", "Opponent", "DIFF", "AST", "BLK", 
                  "STL", "DRB", "ORB", "PIP", "FBP", "FOUL",
                  "TOV", "TOVP", "FTp", "FGp", "P2p", "P3p")

set.seed(100)
# Get the optimal mtry (the one that gives the lowest error rate)
tuning_results <- tuneRF(train[, RFpredictors], train$WL, 
                         ntreeTry=500, stepFactor=1.5, 
                         improve=0.01, trace=TRUE, plot=TRUE)

optimal_mtry <- tuning_results[as.numeric(which.min(tuning_results[, 2])),1]

# Print the optimal mtry
# print(paste("Optimal mtry:", optimal_mtry))

# Convert the tuning results to a data frame
tuning_results_df <- as.data.frame(tuning_results)

# Create a ggplot
ggplot(tuning_results_df, aes(x = mtry, y = tuning_results[, 2])) +
  geom_line() +
  geom_point() +
  geom_point(aes(x = optimal_mtry, y = min(tuning_results[, 2])), colour = 'red') +
  labs(x = "mtry", y = "Out-of-bag error rate", title = "Tuning mtry in Random Forest") +
  theme_minimal()

# Fit the final random forest model with the optimal mtry
final_rf_model <- randomForest(WL ~ ., train[, c("WL", LRpredictors)], 
                               mtry = optimal_mtry, 
                               ntree = 1000, maxnodes = 50, 
                               nodesize = 3, importance = TRUE)
                               
# Predict on the training set
train$RFpredicted_WL <- predict(final_rf_model, 
                                newdata = train[, LRpredictors])

# Print the confusion matrix for the training set
# print(confusionMatrix(train$RFpredicted_WL, train$WL))

# Predict on the test set
test$RFpredicted_WL <- predict(final_rf_model, newdata = test)

# Print the confusion matrix for the test set
# print(confusionMatrix(test$RFpredicted_WL, test$WL))

# Print the variable importance
importance_scores <- final_rf_model$importance

# Convert the importance scores to a data frame for plotting
importance_df <- data.frame(Variable = rownames(importance_scores),
                            Importance = importance_scores[, "MeanDecreaseAccuracy"])

# Sort the variables by importance
importance_df <- importance_df %>%
  arrange(desc(Importance))

# Plot variable importance
ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Variable", 
       y = "Importance",
       title = "Variable Importance Plot",
       subtitle = "Based on Mean Decrease in Accuracy") +
  theme_minimal()

```

## Results
```{r}
candidate = c("NB","KNN","LR","DT","RF")

pred = data.frame(test$NBpredicted_WL,test$KNNpredicted_WL,
         as.factor(test$LR_WL),test$DTpredicted_WL,
         test$RFpredicted_WL)

for (i in 1:length(candidate)){
  CM = confusionMatrix(pred[,i], test$WL)
  print(paste(candidate[i], "Accuracy=",metrics(CM)[1,2]))
  print(paste(candidate[i], "F1 =",metrics(CM)[2,2]))
}

# Create a data frame with the model performance measures
performance_measures <- data.frame(
  Model = c("Naïve Bayes", "K-nearest Neighbors", "Logistic Regression", "Decision Tree", "Random Forest"),
  Training_Accuracy = c("59.68%", "55.28%", "62.32%", "62.09%", "72.47%"),
  Test_F1_score = c("50.44%", "39.26%", "55.52%", "32.62%", "39.84%"),
  Test_Accuracy = c("56.87% (0.54, 0.60)", "52.54% (0.50, 0.55)", "55.35% (0.53, 0.58)", "52.32% (0.50, 0.55)", "53.23% (0.50, 0.56)")
)

# Use kable to create the table
kable(performance_measures, align = c('l', 'c', 'c', 'c'))


# Extract coefficients
coefficients <- coef(bw_model)

# Print the coefficients
print(coefficients)

```

## NBA ranking
```{r}
# Define the data
# data extracted from https://www.basketball-reference.com/leagues/
rk <- read.csv("ranking.csv")

nba_rankings <- data.frame(
  Team = rk$Team,
  Season = c(rep('2018-2019', 30), 
             rep('2019-2020', 30), 
             rep('2020-2021', 30),
             rep('2021-2022', 30)),  
  MOV = c(rk$MOV2019,
          rk$MOV2020,
          rk$MOV2021,
          rk$MOV2022)  
  # Fill in the rest of the MOVs
)

# Create the plot
ggplot(nba_rankings, aes(x = reorder(Team, -MOV), y = MOV, fill = Season)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(x = "Team", y = "Average Margin of Victory (MOV)", 
       title = "NBA Team Rankings (2018-2022)", fill = "Season") +
  theme_minimal()

```






















